---
title: "SSRI Revision - TF analyses"
author: "Susmita Malwade"
date: "2026-02-09"
output: 
  html_document:
    code_folding: "hide"
    df_print: paged
    toc: yes
    theme: cerulean
    toc_float:
      collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r lib, include=FALSE}
library(dplyr)
library(glmnet)
library(Matrix)
library(Seurat)
library(dorothea)
library(viper)
library(Seurat)
library(ggplot2)
library(ggrepel)
library(scales)
library(MASS)
library(boot)
```

## SSRI Revision

We'll try to build one mechanistic argument:

Drug → ARID1B state → TF regulatory programs → ASD-risk genes

ARID1B is a chromatin regulator, not a Transcritption Factor. ARID1B does not act via a regulon but via regulatory context gating. Therefore, we model TF programs as the functional readouts of ARID1B state.


```{r load, include=FALSE}
data(dorothea_hs, package="dorothea")
#regulons_df <- dorothea_hs %>% filter(confidence %in% c("A","B","C","D","E")) %>% select(tf, target, mor)
regulons_df <- readRDS("/Users/susmita.malwade/regulons_df.rds")
regulons_list <- dorothea::df2regulon(regulons_df)
seurat_obj <- readRDS("/Volumes/projects/C3_Sellgren_lab/Lab Members/Susmita/Internal data/CMV & SSRI/scRNAseq/SSRI/SSRI/sertraline/ssri_recent.rds")
DefaultAssay(seurat_obj) <- "RNA"
expr_mat <- GetAssayData(seurat_obj, slot="data")
targets <- unique(unlist(lapply(regulons_list, function(x) names(x$tfmode))))
expr_mat <- expr_mat[intersect(rownames(expr_mat), targets), ]
expr_mat <- as.matrix(expr_mat)
storage.mode(expr_mat) <- "numeric"
expr_mat[is.na(expr_mat)] <- 0
tf_activities <- readRDS("/Users/susmita.malwade/ssri_tf_activities.rds")
seurat_obj[["TF_activity"]] <- CreateAssayObject(tf_activities)

```

### Relationship between TF Activity and DrugScore (proxy for phenotype)

TF activity scores were inferred per cell using curated human regulons from DoRothEA database, and analyses were performed at the donor pseudobulk level (mean activity per donor × condition) to ensure biological replication. A drug-response transcriptional phenotype (DrugScore) was defined as the difference between drug-upregulated and drug-downregulated gene modules.

Lets show that the drug effect is real at the biological replicate level.
```{r pb, include=FALSE}
## ----------------------------
## 1) Build pseudobulk matrices
## ----------------------------

ct <- "RG"  # change as needed

cells_ct <- WhichCells(seurat_obj, expression = new_types == ct)
seurat_ct <- subset(seurat_obj, cells = cells_ct)

de_all <- openxlsx::read.xlsx("/Volumes/projects/C3_Sellgren_lab/Lab Members/Susmita/Internal data/CMV & SSRI/scRNAseq/SSRI/SSRI/sertraline/201024_SSRI_pseudobulk_results_all.xlsx")

de_ct <- de_all %>% filter(celltype == ct, padj < 0.05)
up_genes   <- de_ct %>% filter(log_fc > 0) %>% arrange(desc(log_fc)) %>% pull(feature) %>% head(200)
down_genes <- de_ct %>% filter(log_fc < 0) %>% arrange(log_fc) %>% pull(feature) %>% head(200)
DefaultAssay(seurat_ct) <- "RNA"
seurat_ct <- AddModuleScore(seurat_ct, list(up_genes), name="DrugUp")
seurat_ct <- AddModuleScore(seurat_ct, list(down_genes), name="DrugDown")
seurat_ct$DrugScore <- seurat_ct$DrugUp1 - seurat_ct$DrugDown1

tf_mat <- GetAssayData(seurat_ct, assay="TF_activity", slot="data")  # TFs x cells
stopifnot(all(colnames(tf_mat) %in% colnames(seurat_ct)))

meta <- data.frame(
  cell      = colnames(seurat_ct),
  donor     = seurat_ct$assignment,
  condition = seurat_ct$Condition,
  DrugScore = seurat_ct$DrugScore,
  stringsAsFactors = FALSE
)

# align meta to tf_mat columns
meta <- meta[match(colnames(tf_mat), meta$cell), ]
stopifnot(all(meta$cell == colnames(tf_mat)))

# pseudobulk DrugScore (donor-condition)
pb_score <- meta %>%
  group_by(donor, condition) %>%
  summarize(DrugScore = mean(DrugScore), n_cells = dplyr::n(), .groups="drop")

# pseudobulk TF activity (donor-condition)
groups <- interaction(meta$donor, meta$condition, drop=TRUE)

pb_tf_mat <- sapply(split(seq_len(ncol(tf_mat)), groups), function(idx){
  rowMeans(tf_mat[, idx, drop=FALSE])
})
pb_tf_mat <- t(pb_tf_mat)                  # samples x TFs
colnames(pb_tf_mat) <- rownames(tf_mat)    # TF names as columns

pb_tf <- as.data.frame(pb_tf_mat)
pb_tf$group <- rownames(pb_tf)

pb_score$group <- interaction(pb_score$donor, pb_score$condition, drop=TRUE)

pb_df <- left_join(pb_score, pb_tf, by="group")
rownames(pb_df) <- pb_df$group
pb_df$group <- NULL

# predictors/outcome
X_pb <- scale(as.matrix(pb_df[, rownames(tf_mat)]))  # 8 x 1333
y_pb <- as.numeric(scale(pb_df$DrugScore))

# donor fold id for LODO CV (critical with n=8)
foldid <- as.numeric(as.factor(pb_df$donor))


## ----------------------------
## 2) Fig1: Donor-level drug effect on DrugScore
## ----------------------------
# (Paired lines; p-value paired t-test—interpret cautiously with 4 donors)
p1 <- ggplot(pb_score, aes(condition, DrugScore, group=donor)) +
  geom_line(color="grey70", linewidth=0.8) +
  geom_point(aes(color=donor), size=3) +
  #scale_color_manual(values=c("ctrl"="#4DBBD5","sert"="#E64B35")) +
  theme_classic(base_size=14) +
  labs(title="Drug exposure induces a transcriptional state shift",
       y="DrugScore (pseudobulk)", x="")
```

```{r, fig.align='center'}
p1
```

```{r, include=FALSE}
t.test(DrugScore ~ condition, data=pb_score, paired=TRUE)
```

Drug induces a consistent transcriptional state change across biological replicates. DrugScore explains the drug response (expected since its derived from DEGs.) (sanity check that DrugScore is good to go ahead with). 

#### TF activity predicts DrugScore in RG

We model DrugScore as a function of TF activity using elastic net regression with leave-one-donor-out cross-validation (LODO-CV) (for stats; better than bootstrap) to identify TF programs that robustly predict the drug-induced transcriptional state. TF driver strength was quantified as the cross-validated regression coefficient, and stability was defined as the fraction of LODO folds in which a TF retained a nonzero coefficient.

```{r TF activity, include=FALSE}
## ----------------------------
## 3) Fig2: TF activity predicts DrugScore (LODO CV performance)
## ----------------------------
fit_base <- cv.glmnet(
  x = X_pb, y = y_pb,
  alpha = 0.5,
  foldid = foldid,
  standardize = FALSE
)

pred_cv <- as.numeric(predict(fit_base, newx=X_pb, s="lambda.min"))
cv_r2 <- summary(lm(y_pb ~ pred_cv))$r.squared

p2 <- ggplot(data.frame(obs=y_pb, pred=pred_cv, donor=pb_df$donor, condition=pb_df$condition),
             aes(x=obs, y=pred, shape=condition)) +
  geom_point(size=3) +
  geom_smooth(method="lm", se=FALSE, color="black") +
  theme_classic(base_size=14) +
  labs(title=paste0("TF activity predicts drug-response state (LODO-CV R²=", round(cv_r2, 2), ")"),
       x="Observed DrugScore (scaled)", y="Predicted DrugScore (CV)")
```

```{r, fig.align='center', warning=FALSE}
p2
```

You see two clusters (ctrl vs sert) almost perfectly separated.

### ARID1B's contribution to the phenotype

To test whether ARID1B modulates this regulatory landscape, we'll first treat pseudobulk ARID1B expression as a proxy for chromatin state and check how much of the variance in DrugScore is explained by ARID1B. 
```{r ARID1B, echo=FALSE, warning=FALSE}

## ----------------------------
## 4) Fig3: ARID1B alone does not add explanatory power beyond drug (pseudobulk ΔR²)
## ----------------------------
rna_mat <- GetAssayData(seurat_ct, assay="RNA", slot="data")  # genes x cells
stopifnot("ARID1B" %in% rownames(rna_mat))

arid1b_pb <- meta %>%
  group_by(donor, condition) %>%
  summarize(ARID1B = mean(rna_mat["ARID1B", cell]), .groups="drop")

pb_df2 <- left_join(pb_df, arid1b_pb, by=c("donor","condition"))
pb_df2$ARID1B <- as.numeric(scale(pb_df2$ARID1B))
pb_df2$drug_bin <- ifelse(pb_df2$condition == "sert", 1, 0)

fit_r2_drug <- lm(DrugScore ~ drug_bin, data=pb_df2)
fit_r2_both <- lm(DrugScore ~ drug_bin + ARID1B, data=pb_df2)

r2_drug <- summary(fit_r2_drug)$r.squared
r2_both <- summary(fit_r2_both)$r.squared
delta_r2 <- r2_both - r2_drug
r2_drug
print(paste0("Variance explained by sertraline = ", r2_drug*100, "%"))
```

```{r, echo=FALSE}
print(paste0("Variance explained by sertraline and ARID1B = ", r2_both*100, "%"))
```

```{r, echo=FALSE}
print(paste0("Variance explained by ARID1B alone = ", delta_r2*100, "%"))
```
ARID1B alone does not explain the phenotype beyond the drug exposure. 

#### ARID1B gating test: residualize TF activities vs ARID1B

Next, we'll remove TF variation aligned with ARID1B by regressing each TF activity on ARID1B and using residuals as ARID1B-independent TF activity. Re-fitting the DrugScore model with ARID1B-residualized TF activities yields post-removal coefficients, and ARID1B dependence was defined as the reduction in absolute TF coefficient magnitude after ARID1B-associated variation was removed. This constitutes an in silico conditional perturbation testing whether TF programs drive the drug phenotype within an ARID1B-associated regulatory context.

```{r resid, include=FALSE}
## ----------------------------
## 5) ARID1B gating test: residualize TF activities vs ARID1B, refit TF→DrugScore
##    Use CV coefficients as the driver metric (more defensible than covariance KO)
## ----------------------------
coefs_base <- as.numeric(coef(fit_base, s="lambda.min"))[-1]
names(coefs_base) <- colnames(X_pb)

arid <- pb_df2$ARID1B
X_resid <- apply(X_pb, 2, function(v) resid(lm(v ~ arid)))
X_resid <- scale(as.matrix(X_resid))
colnames(X_resid) <- colnames(X_pb)

fit_resid <- cv.glmnet(X_resid, y_pb, alpha=0.5, foldid=foldid, standardize=FALSE)

coefs_resid <- as.numeric(coef(fit_resid, s="lambda.min"))[-1]
names(coefs_resid) <- colnames(X_resid)

results_pb <- data.frame(
  TF = colnames(X_pb),
  coef_base  = coefs_base[colnames(X_pb)],
  coef_resid = coefs_resid[colnames(X_pb)]
)

results_pb$dependence  <- abs(results_pb$coef_base) - abs(results_pb$coef_resid)
results_pb$drug_driver <- results_pb$coef_base

unique_donors <- unique(pb_df$donor)
coef_lodo <- matrix(0, nrow=length(unique_donors), ncol=ncol(X_pb))
colnames(coef_lodo) <- colnames(X_pb)

for(i in seq_along(unique_donors)){
  keep <- pb_df$donor != unique_donors[i]
  fit_i <- cv.glmnet(X_pb[keep,], y_pb[keep],
                     alpha=0.5,
                     foldid=as.numeric(as.factor(pb_df$donor[keep])),
                     standardize=FALSE)
  coef_lodo[i,] <- as.numeric(coef(fit_i, s="lambda.min"))[-1] != 0
}

results_pb$sel_lodo <- colMeans(coef_lodo)

SFARI <- read.csv("https://raw.githubusercontent.com/mmkim1210/gene-list/main/list/SFARI-Gene_genes_01-13-2021release_02-05-2021export.csv")
SFARI <- filter(SFARI, `gene.score` == 1 | syndromic == 1)
nrow(SFARI)  # 328 genes
risk_genes <- SFARI$gene.symbol

bg_genes <- rownames(GetAssayData(seurat_ct, assay="RNA", slot="data"))
risk_genes_bg <- intersect(risk_genes, bg_genes)
reg_df <- regulons_df %>% filter(target %in% bg_genes)
enrich_df <- lapply(unique(reg_df$tf), function(tf){
targets <- unique(reg_df$target[reg_df$tf==tf])
x <- length(intersect(targets, risk_genes_bg))
M <- length(targets); N <- length(bg_genes); K <- length(risk_genes_bg)
p <- phyper(x-1, M, N-M, K, lower.tail=FALSE)
data.frame(TF=tf, overlap=x, regulon_size=M, pval=p)
}) %>% bind_rows()
enrich_df$padj <- p.adjust(enrich_df$pval, "BH")
enrich_df$neglog10_padj <- -log10(enrich_df$padj + 1e-300)

results_pb <- results_pb %>%
    dplyr::left_join(
        dplyr::select(enrich_df, TF, regulon_size, overlap, padj, neglog10_padj),
        by = "TF"
    )

results_pb <- results_pb %>%
    mutate(
        loss_frac = ifelse(abs(coef_base) > 0,
                           dependence / abs(coef_base),
                           0)
    )

p4 <- ggplot(results_pb, aes(x=drug_driver, y=dependence)) +
  geom_hline(yintercept=0, linetype=2, color="grey60") +
  geom_vline(xintercept=0, linetype=2, color="grey60") +
  geom_point(aes(color=sel_lodo, size = regulon_size), alpha=0.85) +
  scale_color_viridis_c() +
  geom_text_repel(aes(label=TF), max.overlaps=25) +
  theme_classic(base_size=14) +
  labs(title="ARID1B gating of drug-driving TF programs",
       x="TF driver coefficient (DrugScore ~ TFs; CV)",
       y="ARID1B dependence: |coef_base| − |coef_resid|",
       color="LODO stability", size = "Regulon size")
```

```{r, fig.align='center', fig.width=8, fig.height=8, warning=FALSE}
p4
```


Axes:
X = TF driving strength
| Positive - TF program contributes to drug-induced transcriptional reprogramming |
| Negative - TF program maintains pre-drug or alternative cell-state identity     |


Y = how much that driving strength disappears when ARID1B-linked variation is removed

So upper-right TFs = drug drivers that depend on ARID1B state.

Stability - Values near 1 → donor-robust driver. Values near 0 → donor-specific.

#### Risk gene enrichment (per TF regulon)

Disease relevance is assessed by testing enrichment of ASD-risk genes (SFARI) within TF regulons using hypergeometric tests with Benjamini–Hochberg correction. 

```{r, include=FALSE}
## ----------------------------
## 6) Risk gene enrichment (per TF regulon) + overlay on gating plot
## ----------------------------

plot_df <- results_pb %>%
  mutate(
    neglog10_padj = ifelse(is.na(neglog10_padj), 0, neglog10_padj),
    overlap       = ifelse(is.na(overlap), 0, overlap),
    risk_hit      = !is.na(padj) & padj < 0.05
  )


p5 <- ggplot(plot_df, aes(x=drug_driver, y=dependence)) +
    geom_hline(yintercept=0, linetype=2, color="grey60") +
    geom_vline(xintercept=0, linetype=2, color="grey60") +
    geom_point(aes(size=regulon_size, shape=risk_hit, color=overlap), alpha=0.85) + scale_color_viridis_c() +
    geom_text_repel(aes(label=TF), max.overlaps=25) +
    theme_classic(base_size=14) +
    labs(title="ARID1B-dependent TF programs enriched for disorder-risk genes",
         x="TF driver coefficient (DrugScore ~ TFs; CV)",
         y="ARID1B dependence: |coef_base| − |coef_resid|",
         size="Regulon size",
         shape="Risk-enriched regulon",
         color="Risk enrichment (overlap)")
```

```{r, fig.align='center', fig.width=8, fig.height=8, warning=FALSE}
p5
```


Technical check: Do TF programs predict the drug-response state in donors not used to fit the model? This is leave-one-donor-out cross-validation (LODO CV) at the sample level.
```{r, include=FALSE}
## ----------------------------
## 7) Statistical tests at the TF-set level (robust with small N)
## ----------------------------
#
  
foldid <- as.numeric(as.factor(pb_df$donor))

pred_cv <- rep(NA_real_, length(y_pb))

for(d in unique(foldid)){
  train <- foldid != d
  test  <- foldid == d
  
  fit_cv <- cv.glmnet(X_pb[train,], y_pb[train],
                      alpha=0.5,
                      foldid=foldid[train],
                      standardize=FALSE)
  
  pred_cv[test] <- as.numeric(
      predict(fit_cv, X_pb[test,], s="lambda.min")
  )
}
cv_r2 <- cor(pred_cv, y_pb)^2
cv_r2
```

TF programs explain ~30% (R-square) of donor-to-donor variation in the drug-response transcriptional state.

✔ Not overfitting and not donor-specific noise

Q. Are risk-enriched TFs overrepresented among drug drivers?

```{r, echo=FALSE, warning=FALSE}
drug_drivers <- results_pb %>%
  filter(sel_lodo > 0.5) %>%      # donor-robust drivers
  pull(TF)

risk_tf <- enrich_df %>% filter(padj < 0.05) %>% pull(TF)

mat_drug_risk <- matrix(c(
  length(intersect(drug_drivers, risk_tf)),
  length(drug_drivers) - length(intersect(drug_drivers, risk_tf)),
  length(risk_tf) - length(intersect(drug_drivers, risk_tf)),
  length(setdiff(results_pb$TF, union(drug_drivers, risk_tf)))
), nrow=2)

fisher_drug <- fisher.test(mat_drug_risk)
fisher_drug

```
Risk-enriched TFs not overrepresented among drug drivers (Fisher p = 1). The TF programs that drive the drug transcriptional phenotype are not specifically the ones whose regulons are enriched for disease-risk genes.

Q. Are ARID1B-dependent TFs enriched for risk-gene regulons?
```{r, echo=FALSE, warning=FALSE}
plot_df2 <- results_pb %>%
  mutate(risk_hit = padj < 0.05)

wilcox.test(dependence ~ risk_hit, data=plot_df2)

```
This tests whether risk-regulating TFs have higher ARID1B dependence.
ARID1B-dependent TF programs are enriched for risk-gene regulons (Wilcoxon p = 0.0027)

Overrepresentation of risk-enriched regulons among drug-driving TFs and among ARID1B-dependent TFs was tested using Fisher’s exact test and wilcoxon test respectively. 


Overall, we see that sertraline response is broad and ARID1B-dependent part of that response is disorder-relevant. 


Summary: TF activity predicted the drug-response transcriptional state across donors (LODO-CV R² = 0.30). TF programs driving the drug phenotype were not enriched for disorder-risk gene regulators. In contrast, TF programs whose activity depended more strongly on ARID1B-associated state showed significantly greater enrichment for risk-gene regulons (Wilcoxon p = 0.0027), indicating that ARID1B-linked regulation preferentially modulates disorder-relevant transcriptional networks within the broader drug response.

### Directionality: ARID1B-high vs ARID1B-low TF regimes

To determine whether transcription factor (TF) programs were preferentially associated with high versus low ARID1B expression states, we modeled pseudobulk TF activity as a function of ARID1B expression while adjusting for drug condition. For each TF, a linear regression model was fit: TF activity∼ARID1B expression+drug condition , using donor-level pseudobulk profiles. The regression coefficient corresponding to ARID1B expression (β) represents the partial association between TF activity and ARID1B, independent of drug effects. TFs with positive β values were classified as ARID1B-high–associated, indicating higher activity when ARID1B expression is elevated, whereas TFs with negative β values were classified as ARID1B-low–associated, indicating greater activity in low-ARID1B states. These directionally annotated TF programs were subsequently integrated with drug-response driver and risk-gene enrichment analyses.

```{r dir, include=FALSE}
## ----------------------------
## 8) Directionality: ARID1B-high vs ARID1B-low TF regimes
##    Use simple partial association slopes (TF ~ ARID1B + drug),
##    then merge with driver + dependence + risk.
## ----------------------------
drug_bin <- pb_df2$drug_bin

assoc_beta <- apply(X_pb, 2, function(tf){
  coef(lm(tf ~ arid + drug_bin))["arid"]
})

dir_df <- data.frame(TF=names(assoc_beta), beta=as.numeric(assoc_beta)) %>%
  mutate(arid_dir = ifelse(beta > 0, "ARID1B_high_assoc", "ARID1B_low_assoc"))

arid_tf_full <- plot_df %>%
  inner_join(dir_df, by="TF") %>%
  mutate(
    label3 = ifelse(
      risk_hit &
        dependence > quantile(dependence, 0.85, na.rm=TRUE) &
        abs(drug_driver) > quantile(abs(drug_driver), 0.85, na.rm=TRUE) &
        abs(beta) > quantile(abs(beta), 0.8, na.rm=TRUE),
      TF, NA
    )
  )

p6 <- ggplot(arid_tf_full,
             aes(x = beta,
                 y = drug_driver,
                 color = arid_dir,
                 size = dependence)) +

    geom_vline(xintercept = 0, linetype = 2, color = "grey60") +
    geom_hline(yintercept = 0, linetype = 2, color = "grey60") +

    geom_point(alpha = 0.85) +

    geom_text_repel(
        aes(label = ifelse(dependence >= 0.01, TF, NA)),
        max.overlaps = 25
    ) +

    theme_classic(base_size = 14) +

    labs(
        title = "Directionality of ARID1B-linked TF programs (partial association)",
        x = "Partial association with ARID1B (TF ~ ARID1B + drug)",
        y = "Drug-driving TF coefficient (CV)",
        color = "ARID1B linkage",
        size = "ARID1B dependence"
    )

```
```{r, fig.align='center', fig.width=8, fig.height=8, warning=FALSE}
p6
```

We assess convergence across three independent axes: drug-response relevance, ARID1B dependence, and genetic risk enrichment. TF programs that ranked highly across all axes represent regulatory modules where ARID1B-associated chromatin state gates drug-responsive transcriptional programs controlling disease-risk genes.

Top 20 convergent TF programs: 

```{r, warning=FALSE, echo=FALSE}
plot_df2 <- plot_df %>%
    mutate(
        drug_strength = abs(coef_base),
        arid_dependence = dependence,
        risk_signal = neglog10_padj
    )

plot_df2 <- plot_df2 %>%
    mutate(
        z_drug  = scale(drug_strength)[,1],
        z_arid  = scale(arid_dependence)[,1],
        z_risk  = scale(risk_signal)[,1],
        convergence_score = z_drug + z_arid + z_risk
    )

top_convergent <- plot_df2 %>%
    dplyr::arrange(desc(convergence_score)) %>%
    dplyr::select(TF, drug_strength, arid_dependence, risk_signal, convergence_score)

head(top_convergent, 20)
```

```{r, include=FALSE}
p7<- ggplot(plot_df2, aes(x = drug_strength, y = arid_dependence)) +
    geom_point(aes(size = risk_signal, color = convergence_score), alpha = 0.85) +
    geom_text_repel(
        data = subset(plot_df2, convergence_score > quantile(convergence_score, 0.95)),
        aes(label = TF),
        max.overlaps = 25
    ) +
    scale_color_viridis_c(option = "magma") +
    theme_bw() +
    labs(
        x = "|Drug-driving TF coefficient|",
        y = "ARID1B dependence",
        size = "Risk enrichment (−log10 FDR)",
        color = "Multi-layer convergence",
        title = "Convergent TF programs linking drug response, ARID1B state, and genetic risk"
    )
```

```{r, fig.align='center', warning=FALSE}
p7
```

Visualize the TF network where edges = shared target genes

```{r, include=FALSE}
top_tfs <- plot_df2 %>%
    arrange(desc(convergence_score)) %>%
    slice(1:25) %>%          # top 25 convergent TFs
    pull(TF)

reg_sub <- regulons_df %>%
    filter(tf %in% top_tfs)

tf_targets <- split(reg_sub$target, reg_sub$tf)

tf_pairs <- combn(names(tf_targets), 2, simplify = FALSE)

edge_df <- lapply(tf_pairs, function(pair){
    t1 <- pair[1]
    t2 <- pair[2]
    g1 <- tf_targets[[t1]]
    g2 <- tf_targets[[t2]]

    inter <- length(intersect(g1, g2))
    union <- length(union(g1, g2))

    jaccard <- inter / union

    data.frame(from=t1, to=t2, weight=jaccard)
}) %>% bind_rows()

edge_df <- edge_df %>% filter(weight > 0.05)  # remove weak overlaps

node_df <- plot_df2 %>%
    dplyr::filter(TF %in% top_tfs) %>%
    dplyr::select(TF, drug_strength, arid_dependence, risk_signal, risk_hit)

library(igraph)
library(ggraph)

g <- graph_from_data_frame(edge_df, vertices=node_df, directed=FALSE)

```
```{r}
ggraph(g, layout="fr") +
    geom_edge_link(aes(width=weight), alpha=0.4, color="grey60") +
    geom_node_point(aes(
        size = drug_strength,
        fill = arid_dependence,
        shape = risk_hit
    ), color="black", stroke=0.8) +
    scale_shape_manual(values=c(21, 24)) +  # circle & triangle with fill
    scale_fill_viridis_c(option="plasma") +
    scale_size_continuous(range=c(4,12)) +
    geom_node_text(aes(label=name), repel=TRUE, size=4) +
    theme_void()
```



Bright nodes = ARID1B-gated TF programs
Large nodes = drug-driving TFs
Triangles = disease-relevant TFs

Clusters of large + bright + triangles =
ARID1B-dependent regulatory modules controlling disease-risk genes

Lets extract the target genes of ARID1B-associated TFs and the enriched risk genes.

```{r, echo=FALSE, warning=FALSE}
arid_split <- arid_tf_full[which(arid_tf_full$TF %in% top_tfs),]

reg_sub <- regulons_df %>%
    filter(tf %in% arid_split$TF)

reg_risk <- reg_sub %>%
    filter(target %in% risk_genes_bg) %>%
    left_join(arid_split, by=c("tf"="TF"))

risk_summary <- reg_risk %>%
    group_by(arid_dir, tf) %>%
    summarize(
        risk_gene_count = n(),
        risk_genes = paste(unique(target), collapse=", ")
    ) %>%
    arrange(arid_dir, desc(risk_gene_count))

risk_summary
```


### Save results
```{r, include=FALSE}
library(openxlsx)

tf_master <- plot_df %>%
    mutate(
        TF = as.character(TF),
        drug_driver = coef_base,
        arid_dependence = dependence,
        arid_loss_fraction = loss_frac,
        lodo_stability = sel_lodo,
        risk_enrichment_score = neglog10_padj
    ) %>%
    dplyr::select(
        TF,
        drug_driver,
        coef_resid,
        arid_dependence,
        arid_loss_fraction,
        lodo_stability,
        regulon_size,
        overlap,
        risk_enrichment_score,
        padj,
        risk_hit
    )

write.xlsx(tf_master,
           "TF_master_results.xlsx",
           overwrite = TRUE)

tf_convergence <- plot_df2 %>%
    dplyr::select(TF, convergence_score, drug_strength, arid_dependence, risk_signal)

write.xlsx(tf_convergence,
           "TF_convergence_scores.xlsx",
           overwrite = TRUE)

arid_direction <- arid_tf_full %>%
    dplyr::select(TF, beta, arid_dir, drug_driver, dependence)

write.xlsx(arid_direction,
           "ARID1B_directional_TFs.xlsx",
           overwrite = TRUE)

tf_risk_map <- risk_summary %>%
    ungroup() %>%
    rename(
        TF = tf,
        ARID1B_regime = arid_dir,
        n_risk_genes = risk_gene_count,
        risk_gene_list = risk_genes
    )

write.xlsx(tf_risk_map,
           "TF_regulon_risk_gene_summary.xlsx",
           overwrite = TRUE)

```


TF_master_results.xlsx - All TF-level statistics

| Column                | Meaning                                                        |
| --------------------- | -------------------------------------------------------------- |
| TF                    | Transcription factor                                           |
| drug_driver           | TF effect on DrugScore (drug-response driver strength)         |
| coef_resid            | TF effect after ARID1B removal                                 |
| arid_dependence       | Absolute change in TF coefficient after ARID1B residualization |
| arid_loss_fraction    | Fraction of TF drug-driving signal linked to ARID1B            |
| lodo_stability        | Stability across donors (leave-one-donor-out frequency)        |
| regulon_size          | Number of genes in TF regulon                                  |
| overlap               | # disorder-risk genes in regulon                               |
| risk_enrichment_score | −log10(FDR) enrichment of risk genes                           |
| padj                  | FDR-adjusted enrichment p-value                                |
| risk_hit              | TRUE if regulon significantly enriched                         |


TF_regulon_risk_gene_summary sheet - Risk genes in top tf regulons

### Conclusion

Drug exposure reduces ARID1B expression in RG cells. Rather than acting as a direct transcription factor, ARID1B appears to gate which TF programs dominate the transcriptional landscape. Several TF programs that both drive drug-response transcription and depend on ARID1B state are enriched for disorder-risk genes, suggesting chromatin-state–dependent control of disease-relevant regulatory networks.

### End

